\documentclass{llncs}

\usepackage{amsmath}
\usepackage{amssymb}

\newcommand{\GF}{\mathrm{GF}}
\newcommand{\vdm}{\mathrm{vdm}}
\newcommand{\had}{\mathrm{had}}
\newcommand{\matrices}[2]{\mathcal{M}_{#1 \times #2}[\GF(2^8)]}
\newcommand{\compository}{\mathop{\bigcirc}\displaylimits}
\newcommand{\gf}[1]{\textsf{`#1'}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\w}{\mathrm{w_h}}

\sloppy

\begin{document}

\pagestyle{plain}

\title{The {A\normalsize NUBIS} Block Cipher}

\author{Paulo S.L.M. Barreto\inst{1}\thanks{Co-sponsored by the
Laborat\'{o}rio de Arquitetura e Redes de Computadores (LARC) do
Departamento de Engenharia de Computa\c{c}\~{a}o e Sistemas Digitais da
Escola Polit\'{e}cnica da Universidade de S\~{a}o Paulo (Brazil)}
\and Vincent Rijmen\inst{2}\thanks{F.W.O.
Postdoctoral Researcher, sponsored by the Fund for Scientific
Research -- Flanders (Belgium)}}

\institute{Scopus Tecnologia S.~A. \\
A. Mutinga, 4105 - Pirituba\\
BR--05110--000 S\~{a}o Paulo (SP), Brazil\\
{\tt pbarreto@scopus.com.br} \and
Katholieke Universiteit Leuven, Dept. ESAT,\\
Kard. Mercierlaan 94,\\
B--3001 Heverlee, Belgium \\
{\tt vincent.rijmen@esat.kuleuven.ac.be}
}

\maketitle

\begin{abstract}
\textsc{Anubis} is a 128-bit block cipher that accepts a
variable-length key. The cipher is a uniform
substitution-permutation network whose inverse only differs from
the forward operation in the key schedule. The design of both the
round transformation and the key schedule is based upon the Wide
Trail strategy and permits a wide variety of implementation
tradeoffs.
\end{abstract}

\section{Introduction}

In this document we describe \textsc{Anubis}, a 128-bit block
cipher that accepts a variable-length key.

Although \textsc{Anubis} is not a Feistel cipher, its structure is
designed so that by choosing all round transformation components
to be involutions, the inverse operation of the cipher differs
from the forward operation in the key scheduling only. This
property will allow reducing the required chip area in a hardware
implementation, as well as the code and table size, which can be
important when \textsc{Anubis} is used e.g. in a Java applet.

\textsc{Anubis} was designed according to the Wide Trail
strategy~\cite{joan}. In the Wide Trail strategy, the round
transformation of a block cipher is composed of different
invertible transformations, each with its own functionality and
requirements.  The \emph{linear diffusion layer} ensures that
after a few rounds all the output bits depend on all the input
bits. The \emph{nonlinear layer} ensures that this dependency is
of a complex and nonlinear nature. The \emph{round key addition}
introduces the key material. One of the advantages of the Wide
Trail strategy is that the different components can be specified
quite independently from one another. We follow the Wide Trail
strategy in the design of the key scheduling algorithm as well.

This document is organised as follows. The mathematical
preliminaries and notation employed are described in
section~\ref{preliminaries}. A mathematical description of the
\textsc{Anubis} primitive is given in section~\ref{description}. A
statement of the claimed security properties and expected
security level is made in section~\ref{sec-goals}. An analysis of
the primitive with respect to standard cryptanalytic attacks is
provided in section~\ref{analysis} (a statement that there are no
hidden weaknesses inserted by the designers is explicitly made in
section~\ref{no-hidden-weakness}). Section~\ref{rationale}
contains the design rationale explaining design choices.
Implementation guidelines to avoid implementation weaknesses are
given in section~\ref{implementation}. Estimates of the
computational efficiency in software are provided in
section~\ref{efficiency}. The overall strengths and advantages of
the primitive are listed in section~\ref{advantages}.

\section{Mathematical preliminaries and notation}\label{preliminaries}

We now summarise the mathematical background and notation that
will be used throughout this paper.

\subsection{Finite fields}

The finite field $\GF(2^8)$ will be represented as
$\GF(2)[x]/p(x)$, where $p(x) = x^8 + x^4 + x^3 + x^2 + 1$ is the
first primitive polynomial of degree 8 listed
in~\cite{lidl-niederreiter}.
The polynomial $p(x)$ was chosen so
that $g(x) = x$ is a generator of $\GF(2^8) \setminus \{0\}$.

An element $u = u_7 x^7 + u_6 x^6 + u_5 x^5 + u_4 x^4 + u_3 x^3 +
u_2 x^2 + u_1 x + u_0$ of $\GF(2^8)$ where $u_i \in \GF(2)$ for
all $i = 0, \dots, 7$ will be denoted by the numerical value $u_7
\cdot 2^7 + u_6 \cdot 2^6 + u_5 \cdot 2^5 + u_4 \cdot 2^4 + u_3
\cdot 2^3 + u_2 \cdot 2^2 + u_1 \cdot 2 + u_0$, written in
hexadecimal notation (hexadecimal digits enclosed in quotes). For
instance, the polynomial $u = x^4 + x + 1$ will be represented by
the hexadecimal byte value \gf{13}. By extension, the reduction
polynomial $p(x)$ may be written \gf{11d}.

\subsection{Matrix classes}

$\matrices{m}{n}$ denotes the set of $m \times n$ matrices over
$\GF(2^8)$.

$\vdm_n(a_0, a_1, \dots, a_{m-1})$ denotes the $m \times n$
Vandermonde matrix~\cite{hoffman-kunze} whose second column
consists of elements $a_0, a_1, \dots, a_{m-1}$, i.e.
\[
\vdm_n(a_0, a_1, \dots, a_{m-1}) \equiv \left[\begin{array}{ccccc}
1 & a_0^{} & a_0^2 & \dots & a_0^{n-1} \\
1 & a_1^{} & a_1^2 & \dots & a_1^{n-1} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & a_{m-1}^{} & a_{m-1}^2 & \dots & a_{m-1}^{n-1}
\end{array} \right].
\]
We write simply $\vdm(a_0, \dots, a_{m-1})$ where the number of
columns is not important for the discussion, or clear from the
context.

If $m$ is a power of 2, $\had(a_0, \dots, a_{m-1})$ denotes the
$m \times m$ Hadamard matrix~\cite{walsh} with elements $h_{ij} =
a_{i \oplus j}$.

\subsection{MDS codes}

We provide a few relevant definitions regarding the theory of
linear codes. For a more extensive exposition on the subject we
refer to~\cite{macwilliams-sloane}.

The Hamming distance between two vectors $u$ and $v$ from the
$n$-dimensional vector space $\GF(2^p)^n$ is the number of
coordinates where $u$ and $v$ differ.

The Hamming weight $\w(a)$ of an element $a \in \GF(2^p)^n$ is
the Hamming distance between $a$ and the null vector of
$\GF(2^p)^n$, i.e. the number of nonzero components of $a$.

A \emph{linear $[n, k, d]$ code} over $\GF(2^p)$ is a
$k$-dimensional subspace of the vector space $(\GF(2^p))^n$,
where the Hamming distance between any two distinct subspace
vectors is at least $d$ (and $d$ is the largest number with this
property).

A \emph{generator matrix $G$} for a linear $[n, k, d]$ code
$\mathcal{C}$ is a $k \times n$ matrix whose rows form a basis for
$\mathcal{C}$. A generator matrix is in \emph{echelon} or
\emph{standard} form if it has the form $G = [I_{k \times k} \,
A_{k \times (n-k)}]$, where $I_{k \times k}$ is the identity
matrix of order $k$. We write simply $G = [I \, A]$ omitting the
indices wherever the matrix dimensions are irrelevant for the
discussion, or clear from the context.

Linear $[n,k,d]$ codes obey the {\em Singleton bound}:
\[
d \leqslant n - k + 1.
\]

A code that meets the bound, i.e. $d = n - k + 1$, is called a
\emph{maximal distance separable} (MDS) code.

A linear $[n, k, d]$ code $\mathcal{C}$ with generator matrix $G
= [I_{k \times k} \, A_{k \times (n-k)}]$ is MDS if, and only if,
every square submatrix formed from rows and columns of $A$ is
nonsingular (cf.~\cite{macwilliams-sloane}, chapter~11, \S\,4,
theorem~8).

\subsection{Cryptographic properties}

A product of $m$ distinct Boolean variables is called an $m$-th
order product of the variables. Every Boolean function $f:
\GF(2)^n \rightarrow \GF(2)$ can be written as a sum over
$\GF(2)$ of distinct $m$-order products of its arguments, $0
\leqslant m \leqslant n$; this is called the algebraic normal
form of $f$. The \emph{nonlinear order} of $f$, denoted $\nu(f)$,
is the maximum order of the terms appearing in its algebraic
normal form.

A \emph{linear} Boolean function is a Boolean function of
nonlinear order 1, i.e. its algebraic normal form only involves
isolated arguments. Given $\alpha \in \GF(2)^n$, we denote by
$l_{\alpha}: \GF(2)^n \rightarrow GF(2)$ the linear Boolean
function consisting of the sum of the argument bits selected by
the bits of $\alpha$:
\[
l_{\alpha}(x) = \bigoplus_{i=0}^{n-1} \alpha_i \cdot x_i.
\]

A mapping $S: \GF(2^n) \rightarrow \GF(2^n), x \mapsto S[x]$, is
called a \emph{substitution box}, or S-box for short. An S-box
can also be viewed as a mapping $S: \GF(2)^n \rightarrow
\GF(2)^n$ and therefore described in terms of its component
Boolean functions $s_i: \GF(2)^n \rightarrow \GF(2), 0 \leqslant i
\leqslant n-1$, i.e. $S[x] = (s_0(x), \dots, s_{n-1}(x))$.

The \emph{nonlinear order} of an S-box $S$, denoted $\nu_S$, is
the minimum nonlinear order over all linear combinations of the
components of $S$:
\[
\nu_S = \min_{\alpha \in \GF(2)^n}\{\nu(l_{\alpha} \circ S)\}.
\]

The \emph{difference table} of an S-box $S$ is defined as
\[
e_S(a, b) = \#\{c \in \GF(2^n) | S[c \oplus a] \oplus S[c] = b\}.
\]

The \emph{$\delta$-parameter} of an S-box $S$ is defined as
\[
\delta_S = \dfrac{1}{e_S(0,0)} \cdot \max_{a \neq 0, \, b}{e_S(a,
b)}.
\]

The product $\delta \cdot e_S(0,0)$ is called the
\emph{differential uniformity} of $S$.

The \emph{correlation} $c(f, g)$ between two Boolean functions $f$
and $g$ can be calculated as follows:
\[
c(f,g) = 2^{1-n} \cdot \#\{x | f(x) = f(g)\} - 1.
\]

The \emph{$\lambda$-parameter} of an S-box $S$ is defined as the
maximal value for the correlation between linear functions of
input bits and linear functions of output bits of $S$:
\[
\lambda_S = \max_{(i,j) \neq (0,0)} c(l_i, l_j \circ S).
\]

The \emph{branch number $\B$} of a linear mapping $\theta:
\GF(2^p)^k \rightarrow \GF(2^p)^m$ is defined as
\[
\B(\theta) = \min_{a \neq 0}\{\w(a) + \w(\theta(a))\}.
\]
Given an $[k + m, k, d]$ linear code over $\GF(2^p)$ with
generator matrix $G = [I_{k \times k} \, M_{k \times m}]$, the
linear mapping $\theta: \GF(2^p)^k \rightarrow \GF(2^p)^m$
defined by
\[
\theta(a) = a \cdot M
\]
has branch number $\B(\theta) = d$; if the code is MDS, such a
mapping is called an \emph{optimal diffusion
mapping}~\cite{vincent}.

\subsection{Miscellaneous notation}

Given a sequence of functions $f_m, f_{m+1}, \dots, f_{n-1}, f_n,
\, m \leqslant n$, we use the notation
$\compository_{r=m}^n{\,f_r} \equiv f_m \circ f_{m+1} \circ \dots
\circ f_{n-1} \circ f_n$, and $\compository^{r=n}_m{\,f_r} \equiv
f_n \circ f_{n-1} \circ \dots \circ f_{m+1} \circ f_m$; if $m >
n$, both expressions stand for the identity mapping.

\section{Description of the {A\normalsize NUBIS}
primitive}\label{description}

The \textsc{Anubis} cipher is an iterated
`involutional'\footnote{We explain in section~\ref{s_inverse} what we mean by
an `involutional' block cipher.}
block
cipher that operates on a 128-bit \emph{cipher state}. It uses a
variable-length, $32N$-bit \emph{cipher key} ($4 \leqslant N
\leqslant 10$), and consists of a series of applications of a
key-dependent round transformation to the cipher state.

In the following we will individually define the component
mappings and constants that build up \textsc{Anubis}, then specify
the complete cipher in terms of these components. The definitions
are often parameterised by $N$, as many components are used both
in the round structure and in the key schedule.

\subsection{Input and output}

The cipher state is internally viewed as a matrix in
$\matrices{4}{4}$, and the cipher key as a matrix in
$\matrices{N}{4}$. Therefore, 128-bit data blocks and $32N$-bit
cipher keys (externally represented as byte arrays) must be mapped
to and from the internal matrix format. This is done by function
$\mu: \GF(2^8)^{4N} \rightarrow \matrices{N}{4}$ and its inverse:
\[
\mu(a) = b \,\Leftrightarrow\, b_{ij} = a_{4i + j}, \; 0
\leqslant i \leqslant N-1, \; 0 \leqslant j \leqslant 3.
\]

\subsection{The nonlinear layer $\gamma$}

Function $\gamma: \matrices{N}{4} \rightarrow \matrices{N}{4}$,
$4 \leqslant N \leqslant 10$, consists of the parallel application
of a nonlinear substitution box $S: \GF(2^8) \rightarrow
\GF(2^8), \, x \mapsto S[x]$ to all bytes of the argument
individually:
\[
\gamma(a) = b \,\Leftrightarrow\, b_{ij} = S[a_{ij}], \; 0
\leqslant i \leqslant N-1, \; 0 \leqslant j \leqslant 3.
\]
The substitution box was pseudo-randomly chosen and is listed in
appendix~\ref{sbox-def}. The search criteria are described in
section~\ref{sbox-crit}; one of them imposes that $S$ be an
involution, i.e. $S[S[x]] = x$ for all $x \in \GF(2^8)$.
Therefore, $\gamma$ itself is an involution.

\subsection{The transposition $\tau$}

Mapping $\tau: \matrices{4}{4} \rightarrow \matrices{4}{4}$
simply transposes its argument:
\[
\tau(a) = b \,\Leftrightarrow\, b = a^t \,\Leftrightarrow\, b_{ij}
= a_{ji}, \; 0 \leqslant i, j \leqslant 3.
\]
Clearly $\tau$ is an involution.

\subsection{The linear diffusion layer $\theta$}

The diffusion layer $\theta: \matrices{N}{4} \rightarrow
\matrices{N}{4}$, $4 \leqslant N \leqslant 10$, is a linear
mapping based on the $[8, 4, 5]$ MDS code with generator matrix
$G_H = [I \, H]$ where $H = \had(\gf{01}, \gf{02}, \gf{04},
\gf{06})$, i.e.
\[
H = \left[\begin{array}{cccc} %
\gf{01} & \gf{02} & \gf{04} & \gf{06}\\
\gf{02} & \gf{01} & \gf{06} & \gf{04}\\
\gf{04} & \gf{06} & \gf{01} & \gf{02}\\
\gf{06} & \gf{04} & \gf{02} & \gf{01}
\end{array}\right],
\]
so that
\[
\theta(a) = b \,\Leftrightarrow\, b = a \cdot H.
\]
A simple inspection shows that matrix $H$ is symmetric and
unitary. Therefore, $\theta$ is an involution for $N = 4$.

\subsection{The key addition $\sigma[k]$}

The affine key addition $\sigma[k]: \matrices{N}{4} \rightarrow
\matrices{N}{4}$, $4 \leqslant N \leqslant 10$, consists of the
bitwise addition (exor) of a key matrix $k \in \matrices{N}{4}$:
\[
\sigma[k](a) = b \,\Leftrightarrow\, b_{ij} = a_{ij} \oplus
k_{ij}, \; 0 \leqslant i \leqslant N-1, \; 0 \leqslant j
\leqslant 3.
\]
This mapping is also used to introduce round constants in the key
schedule, and is obviously an involution.

\subsection{The cyclical permutation $\pi$}

Permutation $\pi: \matrices{N}{4} \rightarrow \matrices{N}{4}$,
$4 \leqslant N \leqslant 10$, cyclically shifts each column of
its argument independently, so that column $j$ is shifted
downwards by $j$ positions:
\[
\pi(a) = b \,\Leftrightarrow\, b_{ij} = a_{(i-j) \bmod N, j}, \;
0 \leqslant i \leqslant N-1, \; 0 \leqslant j \leqslant 3.
\]

\subsection{The key extraction $\omega$}

The key extraction function $\omega: \matrices{N}{4} \rightarrow
\matrices{4}{4}$, $4 \leqslant N \leqslant 10$, is a linear
mapping based on the $[N+4, N, 5]$ MDS code with generator matrix
$G_V = [I \, V^t]$, where $V = \vdm_N(\gf{01}, \gf{02}, \gf{06},
\gf{08})$, i.e.
\[
V = \left[\begin{array}{lllcl} %
\gf{01} & \gf{01} & \gf{01}^{} & \dots & \gf{01}^{}\\
\gf{01} & \gf{02} & \gf{02}^2 & \dots & \gf{02}^{N-1}\\
\gf{01} & \gf{06} & \gf{06}^2 & \dots & \gf{06}^{N-1}\\
\gf{01} & \gf{08} & \gf{08}^2 & \dots & \gf{08}^{N-1}
\end{array}\right],
\]
so that
\[
\omega(a) = b \,\Leftrightarrow\, b = V \cdot a.
\]

\subsection{The round constants $c^r$}\label{round-const}

The $r$-th round constant ($r > 0$) is a matrix $c^r \in
\matrices{N}{4}$, $4 \leqslant N \leqslant 10$, defined as:
\[
\begin{array}{lcll}
c_{0j}^r & = & S[4(r - 1) + j], & 0 \leqslant j \leqslant 3,\\
c_{ij}^r & = & 0, & 1 \leqslant i < N, 0 \leqslant j \leqslant 3.
\end{array}
\]

\subsection{The key schedule}

The key schedule expands the cipher key $K \in \GF(2^8)^{4N}$, $4
\leqslant N \leqslant 10$, onto a sequence of round keys $K^0,
\dots, K^R$, with $K^r \in \matrices{4}{4}$:
\begin{eqnarray*}
\kappa^0 & = & \mu(K),\\
\kappa^r & = & (\sigma[c^r] \circ \theta
\circ \pi \circ \gamma)(\kappa^{r-1}), \;\; r > 0,\\
K^r & = & (\tau \circ \omega \circ \gamma)(\kappa^r), \;\; 0
\leqslant r \leqslant R;
\end{eqnarray*}
the composite mappings $\psi[c^r] \equiv \sigma[c^r] \circ \theta
\circ \pi \circ \gamma$ and $\phi \equiv \tau \circ \omega \circ
\gamma$ are called, respectively, the $r$-th round \emph{key
evolution function} and the \emph{key selection function}. The
initial $\gamma$ applied to compute $K^0$ plays no cryptographic
role and is only kept for simplicity.

\subsection{The complete cipher}

\textsc{Anubis} is defined for the cipher key $K \in
\GF(2^8)^{4N}$ as the transformation $\textsc{Anubis}[K]:
\GF(2^8)^{16} \rightarrow \GF(2^8)^{16}$ given by
\[
\textsc{Anubis}[K] \equiv \mu^{-1} \circ \alpha_R[K^0, \dots,
K^R] \circ \mu,
\]
where
\[
\alpha_R[K^0, \dots, K^R] = \sigma[K^R] \circ \tau \circ \gamma
\circ \left(\compository^{r=R-1}_1{\sigma[K^r] \circ \theta \circ
\tau \circ \gamma}\right) \circ \sigma[K^0].
\]
The standard number of rounds $R$ is defined as $R = 8 + N$ for
$32N$-bit keys, $4 \leqslant N \leqslant 10$. The composite
mapping $\rho[K^r] \equiv \sigma[K^r] \circ \theta \circ \tau
\circ \gamma$ is called the \emph{round function} (for the $r$-th
round), and the related mapping $\rho'[K^R] \equiv \sigma[K^r]
\circ \tau \circ \gamma$ is called the \emph{last round function}.

\subsection{The inverse cipher} \label{s_inverse}

% An inductive proof would be perhaps more elegant, but also more
% obscure -- even for me :-)

We now show that \textsc{Anubis} is an involutional cipher,
in the sense that the only difference between the cipher and
its inverse is in the key schedule. We will need the following lemmas:

\begin{lemma}
$\tau \circ \gamma = \gamma \circ \tau$.
\end{lemma}
\begin{proof}
This follows from the fact that $\tau$ only transposes its
argument without mixing elements, and $\gamma$ only operates on
individual elements, independently of their coordinates.\qed
\end{proof}

\begin{lemma}
$\theta \circ \sigma[K^r] = \sigma[\theta(K^r)] \circ \theta$.
\end{lemma}
\begin{proof}
It suffices to notice that $(\theta \circ \sigma[K^r])(a) =
\theta(K^r \oplus a) = \theta(K^r) \oplus \theta(a) =
(\sigma[\theta(K^r)] \circ \theta)(a)$, for any $a \in
\matrices{N}{4}$.\qed
\end{proof}

We are now ready to state the main property of the inverse
transform $\alpha^{-1}_R[K^0, \dots, K^R]$:

\begin{theorem}
Let $\bar{K}^0 \equiv K^R$, $\bar{K}^R \equiv K^0$, and
$\bar{K}^r \equiv \theta(K^{R-r})$, $0 < r < R$. Then
$\alpha^{-1}_R[K^0, \dots, K^R] = \alpha_R[\bar{K}^0, \dots,
\bar{K}^R]$.
\end{theorem}
\begin{proof}
We start from the definition of $\alpha_R[K^0, \dots, K^R]$:
\[
\alpha_R[K^0, \dots, K^R] = \sigma[K^R] \circ \tau \circ \gamma
\circ \left(\compository^{r=R-1}_1{\sigma[K^r] \circ \theta \circ
\tau \circ \gamma}\right) \circ \sigma[K^0].
\]
Since the component functions are involutions, the inverse
transform is obtained by applying them in reverse order:
\[
\alpha^{-1}_R[K^0, \dots, K^R] = \sigma[K^0] \circ
\left(\compository_{r=1}^{R-1}{\gamma \circ \tau \circ \theta
\circ \sigma[K^r]}\right) \circ \gamma \circ \tau \circ
\sigma[K^R].
\]
The two above lemmas lead to:
\[
\alpha^{-1}_R[K^0, \dots, K^R] = \sigma[K^0] \circ
\left(\compository_{r=1}^{R-1}{\tau \circ \gamma \circ
\sigma[\theta(K^r)] \circ \theta}\right) \circ \tau \circ \gamma
\circ \sigma[K^R].
\]
The associativity of functional composition allows for slightly
changing the grouping of operations:
\[
\alpha^{-1}_R[K^0, \dots, K^R] = \sigma[K^0] \circ \tau \circ
\gamma \circ \left(\compository_{r=1}^{R-1}{\sigma[\theta(K^r)]
\circ \theta \circ \tau \circ \gamma}\right) \circ \sigma[K^R].
\]
Finally, by substituting $\bar{K}^r$ in the above equation, we
arrive at:
\[
\alpha^{-1}_R[K^0, \dots, K^R] = \sigma[\bar{K}^R] \circ \tau
\circ \gamma \circ \left(\compository^{r=R-1}_1{\sigma[\bar{K}^r]
\circ \theta \circ \tau \circ \gamma}\right) \circ
\sigma[\bar{K}^0].
\]
That is, $\alpha^{-1}_R[K^0, \dots, K^R] = \alpha_R[\bar{K}^0,
\dots, \bar{K}^R]$.\qed
\end{proof}

\begin{corollary}
The \textsc{Anubis} cipher has involutional structure, in the sense
that the only difference between the cipher and its inverse is in
the key schedule.
\end{corollary}

\section{Security goals}\label{sec-goals}

In this section, we present the goals we have set for the
security of \textsc{Anubis}. A cryptanalytic attack will be
considered successful by the designers if it demonstrates that a
security goal described herein does not hold.

In order to formulate our goals, some security-related concepts
need to be defined.

\subsection{The set of ciphers for given block length and key
length}

A block cipher of block length $v$ has $V = 2^v$ possible inputs.
If the key length is $u$ it defines a set of $U = 2^u$
permutations over $\{0,1\}^v$. The number of possible
permutations over $\{0,1\}^v$ is $V!$. Hence the number of all
possible block ciphers of dimensions $u$ and $v$ is
\[
((2^v)!)^{(2^u)} = (V!)^U.
\]
For practical values of the dimensions (e.g. $v$ and $u$ above
40), the subset of block ciphers with exploitable weaknesses form
a negligible minority in this set.

\subsection{K-Security}

\begin{definition}[\cite{joan}]
A block cipher is \emph{K-secure} if all possible attack
strategies for it have the same expected work factor and storage
requirements as for the majority of possible block ciphers with
the same dimensions. This must be the case for all possible modes
of access for the adversary (known/chosen/adaptively chosen
plaintext/ciphertext, known/chosen/adaptively chosen key
relations \dots) and for any a priori key distribution.
\end{definition}
K-security is a very strong notion of security. It can easily be
seen that if one of the following weaknesses apply to a cipher,
it cannot be called K-secure:
\begin{itemize}
\item Existence of key-recovering attacks faster than exhaustive
search;
\item Certain symmetry properties in the mapping (e.g.
any complementation property);
\item Occurrence of non-negligible classes of weak keys;
\item Related-key attacks.
\end{itemize}
K-security is essentially a relative measure. It is quite
possible to build a K-secure block cipher with a 5-bit block and
key length. The lack of security offered by such a scheme is due
to its small dimensions, not to the fact that the scheme fails to
meet the requirements imposed by these dimensions. Clearly, the
longer the key, the higher the security requirements.

\subsection{Hermetic block ciphers}

It is possible to imagine ciphers that have certain weaknesses
and still are K-secure. An example of such a weakness would be a
block cipher with a block length larger than the key length and a
single weak key, for which the cipher mapping is linear. The
detection of the usage of the key would take at least a few
encryptions, while checking whether the key is used would only
take a single encryption.

If this cipher would be used for encipherment, this single weak
key would pose no problem. However, used as a component in a
larger scheme, for instance as the compression function of a hash
function, this property could introduce a way to efficiently
generate collisions.

For these reasons we introduce yet another security concept,
denoted by the term \emph{hermetic}.
\begin{definition}[\cite{joan}]
A block cipher is \emph{hermetic} if it does not have weaknesses
that are not present for the majority of block ciphers with the
same block and key length.
\end{definition}
Informally, a block cipher is hermetic if its internal structure
cannot be exploited in any attack.

\subsection{Goal}
For all allowed key lengths, the security goals are that the
\textsc{Anubis} cipher is:
\begin{itemize}
\item K-secure;
\item Hermetic.
\end{itemize}
If \textsc{Anubis} lives up to its goals, the strength against any
known or unknown attacks is as good as can be expected from a
block cipher with the given dimensions.

\subsection{Expected strength}

\textsc{Anubis} is expected, for all key lengths defined, to
behave as good as can be expected from a block cipher with the
given block and key lengths (in the sense of being K-secure and
hermetic).

This implies among other things, the following. The most
efficient key-recovery attack for \textsc{Anubis} is exhaustive
key search. Obtaining information from given plaintext-ciphertext
pairs about other plaintext-ciphertext pairs cannot be done more
efficiently than by determining the key by exhaustive key search.
The expected effort of exhaustive key search depends on the bit
length of the cipher key and is $2^{m-1}$ applications of
\textsc{Anubis} for $m$-bit keys.

The rationale for this is that a considerable safety margin is
taken with respect to all known attacks. We do however realise
that it is impossible to make non-speculative statements on things
unknown.

\section{Analysis}\label{analysis}

\subsection{Differential and linear cryptanalysis}

Due to the Square pattern propagation theorem (cf.~\cite{vincent},
proposition 7.9), for any two different input values it holds that the
number of S-boxes with a different input value in four consecutive
rounds is at least  $\B^2 = 25$. As a consequence, no differential
characteristic over four rounds has probability larger than
$\delta^{\B^2} = (2^{-5})^{25} = 2^{-125}$. Due to the same theorem,
no linear approximation over four rounds has input-output correlation
larger than $\lambda^{\B^2} = (13 \times 2^{-6})^{25} \approx 2^{-57.5}$.
This makes classical
differential or linear attacks, as well as some advanced variants
like differential-linear attacks, very unlikely to succeed for the
full cipher.

\subsection{Truncated differentials}

The concept of truncated differentials was introduced
in~\cite{knudsen-trunc}, and typically applies to ciphers in
which all transformations operate on well aligned data blocks.
Since in \textsc{Anubis} all transformations operate on bytes
rather than individual bits, we investigated its resistance
against truncated differentials. For 6 rounds or more, no attacks
faster than exhaustive key search have been found.

\subsection{Interpolation attacks}

Interpolation attacks~\cite{interpolation} generally depend on the
cipher components (particularly the S-box) having simple
algebraic structures that can be combined to give expressions
with manageable complexity. In such attacks, the attacker
constructs polynomials (or rational expressions) using cipher
input/output pairs; if these polynomials have small degree, only
few cipher input/output pairs are necessary to solve for their
(key-dependent) coefficients. The complicated expression of the
pseudo-randomly generated S-box in $\GF(2^8)$, in combination with
the effect of the diffusion layer, makes these types of attack
infeasible for more than a few rounds.

\subsection{Weak keys}

The weak keys discussed in this subsection are keys that result
in a block cipher mapping with detectable weaknesses. The best
known case of such weak keys are those of IDEA~\cite{joan}.
Typically, this weakness occurs for ciphers in which the
nonlinear operations depend on the actual key value. This is not
the case for \textsc{Anubis}, where keys are applied using exor
and all nonlinearity is in the fixed S-box. In \textsc{Anubis},
there is no restriction on key selection.

\subsection{Related-key cryptanalysis}

Related-key attacks generally rely upon slow diffusion and/or
symmetry in the key schedule. The \textsc{Anubis} key schedule
inherits many properties from the round structure itself, and was
designed to cause fast, nonlinear diffusion of cipher key
differences to the round keys. In particular, the Vandermonde
matrix $V$ in the key selection seems very effective in countering
all kinds of key based attacks known.

For key lengths that are larger than the length
of one round key, it is inevitable that there exist sets of keys
that produce identical values for at least one round key. Of special
interest for a related key attack seems the possibility to find
two different keys with identical values for two consecutive
round keys.

The code defined by $V$ has distance 5 and operates separately on the 4 columns
of $\kappa^r$. This means that two different $\kappa^r$-values can produce an equal
round key $K^r$ only if for all 4 columns it holds that
they differ in either zero or at least 5 bytes. In order to produce equal round
keys in two consecutive rounds $r$ and $r+1$, this requirement has to be fulfilled
for $\kappa^r$ and $\kappa^{r+1}$. The diffusion layer $\theta$ that is employed
in the key evolution function ensures that in total at least 5 columns have to
be different in $\kappa^r$ and $\kappa^{r+1}$. We leave it as an open problem
to determine whether classes of keys can be determined that produce identical
round key values for two or more consecutive rounds. Even so, it is unclear
how such keys could possibly be used successfully in a related key attack.


\subsection{The Square attack and its variants}

In this section we describe an attack first presented
in~\cite{square}, together with its variants~\cite{partial-sum}.
We focus our attention on the $\alpha_R$ transform, as the
occurrences of $\mu$ and its inverse merely change the data
representation. We will denote by $a^r$ the cipher state at the
beginning of the $r$-round (input to $\gamma$), and by $b^r$ the
cipher state at the output of the $\sigma$ key addition in the
$r$-round; these quantities may be indexed to select a particular
byte. For instance, $b^1_{ij}$ is the byte at position $(i, j)$
of the cipher state at the output of round 1.

%Four rounds:
%\[
%\sigma[K^4] \circ \tau \circ \gamma \circ \sigma[K^3] \circ
%\theta \circ \tau \circ \gamma \circ \sigma[K^2] \circ \theta
%\circ \tau \circ \gamma \circ \sigma[K^1] \circ \theta \circ \tau
%\circ \gamma \circ \sigma[K^0]
%\]
%
%Five rounds:
%\[
%\sigma[K^5] \circ \tau \circ \gamma \circ \sigma[K^4] \circ
%\theta \circ \tau \circ \gamma \circ \sigma[K^3] \circ \theta
%\circ \tau \circ \gamma \circ \sigma[K^2] \circ \theta \circ \tau
%\circ \gamma \circ \sigma[K^1] \circ \theta \circ \tau \circ
%\gamma \circ \sigma[K^0]
%\]

\subsubsection{The basic 4-round attack:}
Take a set of 256 plaintexts different from each other in a
single byte (which assumes all possible values), the remaining 15
bytes being constant. After two rounds all 16 bytes of each
cipher state $a^3$ in the set will take every value exactly once.
After three rounds, the exor of all 256 cipher states $a^4$ at
every byte position will be zero.

Consider a ciphertext $b^4 = \tau(\gamma(a^4)) \oplus K^4$;
clearly $a^4 = \gamma(\tau(b^4) \oplus \tau(K^4))$. Now take a
byte from $\tau(b^4)$, guess the matching byte from $\tau(K^4)$
and apply $\gamma$ to the exor of these quantities. Do this for
all 256 ciphertexts in the set and check whether the exor of the
256 results indeed equals zero. If it doesn't, the guessed key
byte is certainly wrong. A few wrong keys (a fraction about
$1/256$ of all keys) may pass this test; repeating it for a second
set of plaintexts leaves only the correct key value with
overwhelming probability.

\subsubsection{Adding a round at the end:}
The 4-round attack can be extended with an extra round at the
end. In this case, the round lacking $\theta$ is the 5th rather
than the 4th. We initially observe that the byte $a^4_{ij}$
depends on the whole row $j$ of $b^4$, which in turn depends on
the whole column $j$ of $b^5$. To obtain it, first guess column
$j$ of $K^5$ and compute row $j$ of $b^4 = \gamma(\tau(b^5 \oplus
K^5))$ (i.e. compute $b^4_{jk} = S[b^5_{kj} \oplus K^5_{kj}]$ for
$k = 0, \dots, 3$). Now take the byte at position $(i, j)$ of
$\tau(\theta(b^4))$, guess the matching byte from
$\tau(\theta(K^4))$ and compute $S$ on the exor of these
quantities, recovering $a^4_{ij}$ and proceeding as in the
4-round attack.

This attack recovers four bytes of the last round key and one byte
of the last round key but one. The remaining bytes can
be obtained in a variety of ways; in the simplest case the process above could
be merely repeated four times, though three applications and a final exhaustive
search for the few missing bytes are often enough, and more efficient.

Overall, $2^{40}$ partial key values must be checked. Since
each set of plaintexts leaves about $1/256$ of wrong keys, the
whole process must be repeated for 5 sets of 256 plaintexts.
However, after testing with the first set of 256 plaintexts,
only $2^{40} \times 2^{-8} = 2^{32}$ partial key values survive,
and only this fraction has to be tested with the second set of
plaintexts. Therefore,
the first check determines the complexity of the attack.
The
attack complexity is therefore $2^{40}$ key guesses
$\times 2^8$ plaintexts
$= 2^{48}$ S-box lookups.

\subsubsection{Adding a round at the beginning:}
The basic idea is to choose a set of 256 plaintexts such that a
single byte of $b^1$ takes all possible values over the set while
the other 15 bytes remain constant, then proceed as with the
5-round attack above. This is achieved by selecting $2^{32}$
plaintexts with one column taking all $2^{32}$ values and the
remaining 12 bytes constant, then guessing the four bytes of
$K^0$ on the same column, and filtering 256 plaintexts that differ
from each other in a single byte on that column.

There is a better way to add a round at the beginning, though.
The $2^{32}$ plaintexts above may be viewed as $2^{24}$ groups of
256 encryptions that differ from each other in a single byte of
$b^1$. Since the exor of the bytes at any position $(i, j)$ of
$a^5$ over each group is zero, the exor at that position over all
$2^{32}$ plaintexts is also zero. Thus, $a^5_{ij}$ is recovered
as in the 5-round attack by guessing column $j$ of $K^6$ plus one
byte at position $(i, j)$ of $\tau(\theta(K^5))$, and the result
is exored over the $2^{32}$ encryptions, checking for zero. This
test is somewhat weaker than the test in the 5-round attack and
may still leave about $1/256$ of wrong keys; therefore it needs
about $6 \times 2^{32}$ chosen plaintexts.
The effort involved is $2^{40}$ key guesses $\times$
$2^{32}$ chosen plaintexts $= 2^{72}$ S-box lookups.

%The effort involved is
%4 columns $\times$ $2^{40}$ key guesses $\times$ $6 \times
%2^{32}$ chosen plaintexts $= 6 \times 2^{74}$ S-box lookups, or
%about $2^{70}$ 7-round encryptions.

\subsubsection{The partial sum improvement:} This is a dynamic
programming technique; it trades computational effort for storage
by reorganising the intermediate computations.

Consider the 6-round attack, where we compute the exor of a byte
$a^5_{ij}$ over the set of encryptions by guessing five key bytes.
Let $k_0, \dots, k_4$ be the guessed key bytes; we may write
\begin{equation}\label{y-byte}
a^5_{ij} = S[\bigoplus_{t=0}^3{S_t[c_t \oplus k_t]} \oplus k_4],
\end{equation}
where the $S_t$ are bijective S-boxes consisting of $S$ followed
by a multiplication by an element from the matrix $H$ used in
$\theta$, and $c_t$ are the ciphertext bytes on column $j$. Let
$x_k = \bigoplus_{t=0}^k{S_t[c_t \oplus k_t]}$. When computing
the exor of the $a^5_{ij}$ bytes, terms with identical values of
$x_k, c_{k+1}, \dots, c_3$ cancel each other. Therefore we speed
up the exor computation by first guessing the values of $k_0$ and
$k_1$ and counting (mod~2) the occurrences of each triple $(x_1,
c_2, c_3)$ over the set of ciphertexts, then guessing $k_2$ and
counting (mod~2) the occurrences of each pair $(x_2, c_3)$ over
the triples $(x_1, c_2, c_3)$, then guessing $k_3$ and counting
$x_3$ over the pairs $(x_2, c_3)$, and finally guessing $k_4$ and
computing the sum over the values of $x_3$.

Notice that we initially guessed $2^{16}$ pairs $(k_0, k_1)$ and
processed $2^{32}$ ciphertexts; for each pair $(k_0, k_1)$ we
guessed $2^8$ values of $k_2$ and processed $2^{24}$ triples
$(x_1, c_2, c_3)$; for each triple $(k_0, k_1, k_2)$ we guessed
$2^8$ values of $k_3$ and processed $2^{16}$ pairs $(x_2, c_3)$;
for each quadruple $(k_0, k_1, k_2, k_3)$ we guessed $2^8$ values
of $k_4$ and processed $2^8$ values of $x_3$. Overall, this
amounts to $2^{48}$ evaluations of equation~\ref{y-byte}. This
effort must be repeated for each of the 6 sets of encryptions used
in the attack, resulting in $6 \times 2^{48}$ S-box lookups, or
about $2^{44}$ 6-round encryptions. The space requirement is
$2^{24} + 2^{16} + 2^{8}$ bits for the counters.

There is an extension of the partial sum attack against 7 rounds
of \textsc{Anubis}~\cite{partial-sum} requiring $2^{128} -
2^{119}$ plaintexts, $2^{64}$ bits of storage and an effort of
about $2^{120}$ encryptions, and an extension to 8 rounds
requiring $2^{128} - 2^{119}$ plaintexts, $2^{104}$ bits of
storage and an effort of about $2^{204}$ encryptions. Although
theoretically interesting, these extensions are really ``on the
edge'': regardless of the cipher being used, an attacker that
manages to convince the key owner to encrypt $99.8\%$ of all
possible plaintexts under the same key could build a dictionary
and decrypt or forge at will an equal fraction of all possible
message blocks, or an arbitrary message block with $99.8\%$ probability --
without ever knowing the key and without the effort of $2^{120}$
(let alone $2^{204}$) extra encryptions.

\subsection{The Gilbert-Minier attack}

The previous attack uses the fact that three rounds of
\textsc{Anubis} can easily be distinguished from a random
permutation. The Gilbert-Minier attack~\cite{gilbert-minier} is
based on a 4-round distinguisher. The attack breaks 7 rounds of
\textsc{Anubis} with complexity: $2^{32}$ guesses for one column
of the first round key $\times$ $2^{16}$ $c$-sets $\times$ 16
encryptions per entry $\times$ $2^{80}$ entries/table $\times$ 2
tables $= 2^{133}$ encryptions (about $2^{140}$ S-box lookups),
plus $2^{32}$ chosen plaintexts. There is a speedup for 128-bit
keys, making the attack marginally faster than exhaustive search,
according to the attack authors.

\subsection{A general extension attack}

Stefan Lucks~\cite{lucks} presents a general extension of any
$n$-round attack; the result is an attack against $(n+1)$ or more
rounds that works for long keys. The idea is simply to guess the
whole $K^{n+1}$ round key and proceed with the $n$-round attack.
Each extra round increases the complexity by a factor $2^{128}$
S-box lookups.

Starting from the 7-round Gilbert-Minier attack, the 8-round
extension costs $2^{140} \times 2^{128}$ S-box lookups, or about
$2^{261}$ 8-round encryptions; therefore it is faster than
exhaustive key search for \textsc{Anubis} keys longer than 256
bits. An extension to 9 rounds against 320-bit keys would require
an 8-round attack of complexity $2^{192}$ encryptions or less, but
none is known.

\subsection{Other attacks}

Attacks based on linear cryptanalysis can sometimes be improved
by using nonlinear approximations \cite{nonl}. However, with the current
state of the art the application of nonlinear approximations seems limited
to the first and/or the last round of a linear approximation. This seems
to be even more so
for ciphers using strongly nonlinear S-boxes, like \textsc{Anubis}.

The impossible differential attack described
in~\cite{biham-keller} can be adapted to work against 5-round
\textsc{Anubis} with $2^{29.5}$ chosen plaintexts and $2^{31}$
steps.

The boomerang attack~\cite{boomerang} benefits from ciphers whose
strength is different for encryption and decryption; this is
hardly the case for \textsc{Anubis}, due to its involutional
structure.

We were not able to find any other method to attack the cipher
faster than exhaustive key search.

\subsection{Designers' statement on the absence of hidden
weaknesses}\label{no-hidden-weakness}

In spite of any analysis, doubts might remain regarding the
presence of trapdoors deliberately introduced in the algorithm.
That is why the NESSIE effort asks for the designers' declaration
on the contrary.

Therefore we, the designers of \textsc{Anubis}, do hereby declare
that there are no hidden weaknesses inserted by us in the
\textsc{Anubis} primitive.

\section{Design rationale}\label{rationale}

\subsection{Self-inverse structure}

Involutional structure is found as part of many cipher designs.
All classical Feistel networks~\cite{feistel} have this property,
as do some more general iterated block ciphers like
IDEA~\cite{idea}. Self-inverse ciphers similar to \textsc{Anubis}
were described and analyzed
in~\cite{youssef-tavares-heys,youssef-mister-tavares}.

The importance of involutional structure resides not only in the
advantages for implementation, but also in the equivalent
security of both encryption and decryption~\cite{knudsen-wagner}.

\subsection{Choice of the substitution box}\label{sbox-crit}

The S-box $S$ was pseudo-randomly chosen to satisfy the following
conditions:
\begin{itemize}
\item $S$ must be an involution, i.e. $S[S[x]] = x$ for all $x \in
\GF(2^8)$.
\item The $\delta$-parameter must not exceed $8 \times 2^{-8}$.
\item The $\lambda$-parameter must not exceed $16 \times 2^{-6}$.
\item The nonlinear order $\nu$ must be maximum, namely, 7.
\end{itemize}

The values of $\delta$ and $\lambda$ are constrained to be no
more than twice the minimum achievable values. The actual
\textsc{Anubis} S-box has $\lambda = 13 \times 2^{-6}$;
experiences showed that involutions with $\delta < 8 \times
2^{-8}$ and $\lambda < 13 \times 2^{-6}$ are extremely rare, as
none was found in a set of over 600 million randomly generated
S-boxes.
%On the other hand, these exact figures are not too difficult to
%achieve: in the same set of random involutions, four were found
%to satisfy those constraints.

The following auxiliary conditions were also imposed to speed up
the S-box search:
\begin{itemize}
\item $S$ must not have any fixed point, i.e. $S[x] \neq x$
for all $x \in \GF(2^8)$.
\item The value of any difference $x \oplus S[x]$ must occur exactly
twice (hence the set of all difference values consists of exactly
128 elements).
\end{itemize}

The absence of fixed points is inspired by the empirical study
reported in section 2.3 of~\cite{youssef-tavares-heys}, where the
strong correlation found between the cryptographic properties and
the number of fixed points of a substitution box suggests
minimising the number of such points. In a more general fashion,
we empirically found that the fraction of random involutions with
good values of $\delta$ and $\lambda$ is increased not only by
avoiding fixed points, but also by minimising the number of
occurrences of any particular difference $x \oplus S[x]$ (or,
equivalently, maximising the number of such differences).
% Perhaps this is due to the maximally contrasting behavior of
% the S-box compared to pure affine mapping?

Finally, the polynomial and rational representations of $S$ over
$\GF(2^8)$ were checked to avoid any obvious algebraic weakness.
The random nature of the search tend to make these
representations as involved as possible.

\subsection{Choice of the diffusion layer}

The actual matrix used in the diffusion layer $\theta$ was
selected by exhaustive search. Although other ciphers of the same
family as \textsc{Anubis} use circulant matrices for this purpose
(cf.~\cite{square,rijndael}), it is not difficult to prove that no
such matrix can be self-inverse. On the other hand, unitary
Hadamard matrices can be easily computed that satisfy the MDS
condition.

The actual choice involves coefficients with the lowest possible
Hamming weight (which is advantageous for hardware
implementations) and lowest possible integer values (which is
important for smart card implementations as discussed in
section~\ref{smart-cards}).

\subsection{On the transposition $\tau$ and the permutation $\pi$}

Though equivalent form the security viewpoint, $\tau$ and $\pi$
have quite different structural properties. The round function
needs an involution to disperse bytes, hence $\pi$ is not
suitable for this task. The key schedule needs a mapping that
keeps the key state dimensions ($N \times 4$) unchanged, hence
$\tau$ is not suitable in general. Therefore both $\tau$ and
$\pi$ are needed in different contexts; for their appointed
purposes, they are perhaps the simplest choices.

\subsection{Structure of the key schedule}

The key scheduling algorithm of \textsc{Anubis} is based on ideas
from~\cite{vincent}. The division into separate layers for
diffusion and confusion is directly taken from the round
structure itself, which ultimately follows the Wide Trail
strategy. The goal is to thwart attacks based on straightforward
exploitation of relationships between round keys. The present
choice also favours component reuse.

\subsection{Choice of the key extraction function}

The Vandermonde structure of the extraction function $\omega$ has
several interesting properties.

First, it is not difficult to obtain MDS Vandermonde matrices by
random search if the number of rows is not too large. The reason
is simple: many submatrices of a Vandermonde matrix $V$ are
themselves Vandermonde matrices (e.g. submatrices built from the
even columns of $V$), while the determinant of many other
submatrices divides the determinant of some Vandermonde matrix
(e.g. submatrices built from consecutive columns of $V$).  Thus,
merely choosing distinct nonzero elements on the second column of
$V$ ensures that many submatrices are nonsingular, and other
simple tests speed up the search even more.

Second, it is possible to choose Vandermonde matrices with
important properties for the key schedule. Thus, it was required
that $\theta \circ \tau \circ \omega$, which is the effective key
extraction function in the inverse cipher, satisfy the same
conditions as $\omega$ itself; therefore, all square submatrices
of $H \cdot V$ are nonsingular. We remark \emph{en passant} that
the occurrence of $\tau$ is motivated by the simple condition it
implies on the product $H \cdot V$ (if $\tau$ were absent, these
matrices would effectively multiply the key state at opposite
sides). The restriction to 320-bit keys comes from the fact that
this condition is no longer satisfied by extensions of $V$ with
more than 10 columns for the actually chosen $V$ coefficients.

Third, multiplication by a Vandermonde matrix can be done quite
efficiently with a fast polynomial evaluation algorithm, as seen
in section~\ref{phi-implementation}. The ease to find suitable
matrices provides freedom in the choice of coefficients, which
can make the polynomial evaluation even more efficient. The
actual coefficients were chosen to have the lowest possible
Hamming weight (which is advantageous for hardware
implementations) while satisfying the condition on $\theta \circ
\tau \circ \omega$ as explained above.

\subsection{Choice of the round constants}

Good round constants should not be equal for all bytes in a state,
and also not equal for all bit positions in a byte. They should
also be different in each round. Making only one row of $c^r$
different from zero makes the round constants simpler. The actual
choice meets these constraints while also reusing an available
component (the S-box itself).

\section{Implementation}\label{implementation}

\textsc{Anubis} can be implemented very efficiently. On different
platforms, different optimisations and tradeoffs are possible. We
make here a few suggestions.

\subsection{32-bit and 64-bit processors} \label{large-processors}

\subsubsection{Implementation of
$\rho$:}\label{rho-implementation}

We suggest the following lookup-table approach.
\begin{eqnarray*}
b &=& (\theta \circ \tau \circ \gamma) (a) \\
 & \Updownarrow & \\
\left[ \begin{array}{cccc}
b_{0,0} & b_{0,1} & b_{0,2} & b_{0,3} \\
b_{1,0} & b_{1,1} & b_{1,2} & b_{1,3} \\
b_{2,0} & b_{2,1} & b_{2,2} & b_{2,3} \\
b_{3,0} & b_{3,1} & b_{3,2} & b_{3,3}
 \end{array} \right]
 &=&
\left[ \begin{array}{cccc}
S[a_{0,0}] & S[a_{1,0}] & S[a_{2,0}] & S[a_{3,0}] \\
S[a_{0,1}] & S[a_{1,1}] & S[a_{2,1}] & S[a_{3,1}] \\
S[a_{0,2}] & S[a_{1,2}] & S[a_{2,2}] & S[a_{3,2}] \\
S[a_{0,3}] & S[a_{1,3}] & S[a_{2,3}] & S[a_{3,3}]
 \end{array} \right]  \cdot
\left[ \begin{array}{cccc}
\gf{01} & \gf{02} & \gf{04} & \gf{06}\\
\gf{02} & \gf{01} & \gf{06} & \gf{04}\\
\gf{04} & \gf{06} & \gf{01} & \gf{02}\\
\gf{06} & \gf{04} & \gf{02} & \gf{01}
\end{array} \right] \\
& \Updownarrow & \\
\left[ \begin{array}{cccc} b_{k,0} & b_{k,1} & b_{k,2} & b_{k,3} \end{array} \right] &=& %
 S[a_{0,k}] \cdot [ \begin{array}{cccc} \gf{01} & \gf{02} & \gf{04} & \gf{06} \end{array} ] \oplus %
 S[a_{1,k}] \cdot [ \begin{array}{cccc} \gf{02} & \gf{01} & \gf{06} & \gf{04} \end{array} ] \\
 & \oplus &
 S[a_{2,k}] \cdot [ \begin{array}{cccc} \gf{04} & \gf{06} & \gf{01} & \gf{02} \end{array} ] \oplus %
 S[a_{3,k}] \cdot [ \begin{array}{cccc} \gf{06} & \gf{04} & \gf{02} & \gf{01} \end{array} ],\\
& 0 & \leqslant k \leqslant 3.
\end{eqnarray*}
Using the following four tables:
\[
\begin{array}{lcl}
T_0[x] = S[x] \cdot [ \begin{array}{cccc} %
\gf{01} & \gf{02} & \gf{04} & \gf{06}
\end{array} ], & \; &
T_1[x] = S[x] \cdot [ \begin{array}{cccc} %
\gf{02} & \gf{01} & \gf{06} & \gf{04}
\end{array} ], \\
T_2[x] = S[x] \cdot [ \begin{array}{cccc} %
\gf{04} & \gf{06} & \gf{01} & \gf{02}
\end{array} ], & &
T_3[x] = S[x] \cdot [ \begin{array}{cccc} %
\gf{06} & \gf{04} & \gf{02} & \gf{01}
\end{array} ],
\end{array}
\]
a row of $b$ can be calculated with four table lookups and three
exor operations; the key addition then completes the evaluation
of $\rho$. The $T$-tables require $4 \times 2^8$ bytes of storage
each. An implementation can use the fact that the corresponding
entries of different $T$-tables are permutations of one another
and save some memory at the expense of introducing extra
permutations at runtime. Usually this decreases the performance
of the implementation.

\subsubsection{Implementation of $\phi$:}\label{phi-implementation}

By definition, $K^r = (\tau \circ \omega \circ \gamma)(\kappa^r)
= (V \cdot \gamma(\kappa^r))^t = \gamma(\kappa^r)^t \cdot V^t$, so
that $K^r_{ij} = \bigoplus_{k=0}^{N-1}{S[\kappa^r_{ki}] \cdot
a_j^k}$, which can be computed by the following algorithm:
\begin{tabbing}margin \= \kill
    \> set $K^r_{ij} = S[\kappa^r_{N-1,\,i}]$; \\
    \> for \= ($k = N - 2$; $k \geqslant 0$; $k\!-\!\!-$) \{ \\
    \>     \> set $K^r_{ij} = S[\kappa^r_{ki}] \,\oplus\, a_j \!\cdot\! K^r_{ij}$; \\
    \> \}
\end{tabbing}
The columns of $K^r_{ij}$ can be computed in parallel:
\begin{tabbing}margin \= \kill %
    \> set $[\begin{array}{cccc}K^r_{i0} & K^r_{i1} & K^r_{i2} & K^r_{i3}\end{array}]
    = S[\kappa^r_{N-1,\,i}] \!\cdot\! [\begin{array}{cccc}\gf{01} & \gf{01} & \gf{01} & \gf{01}\end{array}]$; \\ %
    \> for \= ($k = N - 2$; $k \geqslant 0$; $k\!-\!\!-$) \{ \\ %
    \>     \> set $[\begin{array}{cccc}K^r_{i0} & K^r_{i1} & K^r_{i2} & K^r_{i3}\end{array}]$
    \= $= S[\kappa^r_{ki}] \!\cdot\! [\begin{array}{cccc}\gf{01}& \gf{01} & \gf{01} & \gf{01}\end{array}]$ \\ %
    \>     \>     \> $\oplus \; K^r_{i,0} \!\cdot\! [\begin{array}{cccc}\gf{01} & \gf{00} & \gf{00} & \gf{00}\end{array}]$ \\ %
    \>     \>     \> $\oplus \; K^r_{i,1} \!\cdot\! [\begin{array}{cccc}\gf{00} & \gf{02} & \gf{00} & \gf{00}\end{array}]$ \\ %
    \>     \>     \> $\oplus \; K^r_{i,2} \!\cdot\! [\begin{array}{cccc}\gf{00} & \gf{00} & \gf{06} & \gf{00}\end{array}]$ \\ %
    \>     \>     \> $\oplus \; K^r_{i,3} \!\cdot\! [\begin{array}{cccc}\gf{00} & \gf{00} & \gf{00} & \gf{08}\end{array}]$; \\ %
    \> \}
\end{tabbing}
Using the following two tables:
\[
\begin{array}{lcl}
T_4[x] = S[x] \cdot [\begin{array}{cccc}\gf{01} & \gf{01} & \gf{01} & \gf{01}\end{array}], %
& \; & %
T_5[x] = x    \cdot [\begin{array}{cccc}\gf{01} & \gf{02} & \gf{06} & \gf{08}\end{array}], %
\end{array}
\]
a row of $K^r_{ij}$ can be calculated with $1 + 5(N-1)$ table
lookups, $4(N-1)$ bitwise `and' operations (to properly mask the
results of the $T_5$ lookups), and $4(N-1)$ exor operations.

Alternatively, if enough storage (including processor built-in
cache) is available for the following $N$ tables:
\[
U_k[x] = S[x] \cdot [\begin{array}{cccc}\gf{01}^{} & \gf{02}^k &
\gf{06}^k & \gf{08}^k \end{array}], \; 0 \leqslant k \leqslant
N-1,
\]
then each row $K^r$ can be computed as
$[\begin{array}{cccc}K^r_{i0} & K^r_{i1} & K^r_{i2} &
K^r_{i3}\end{array}] = \bigoplus_{k=0}^{N-1}{U_k[\kappa^r_{ki}]}$
with only $N$ table lookups and $N-1$ exor operations.

\subsubsection{Implementation of $\psi$:}

The same lookup-table approach suggested for $\rho$ can be
employed for $\psi$, reusing the same $T$-tables:
\begin{eqnarray*}
b &=& (\theta \circ \pi \circ \gamma) (a) \\
 & \Updownarrow & \\
 \left[ \begin{array}{cccc} b_{k,0} & b_{k,1} & b_{k,2} & b_{k,3} \end{array} \right] &=& %
 S[a_{k,0}] \cdot [ \begin{array}{cccc} \gf{01} & \gf{02} & \gf{04} & \gf{06} \end{array} ] \\%
 & \oplus &
 S[a_{(k-1)\bmod N,1}] \cdot [ \begin{array}{cccc} \gf{02} & \gf{01} & \gf{06} & \gf{04} \end{array} ] \\
 & \oplus &
 S[a_{(k-2)\bmod N,2}] \cdot [ \begin{array}{cccc} \gf{04} & \gf{06} & \gf{01} & \gf{02} \end{array} ] \\%
 & \oplus &
 S[a_{(k-3)\bmod N,3}] \cdot [ \begin{array}{cccc} \gf{06} & \gf{04} & \gf{02} & \gf{01} \end{array} ],\\
 & 0 & \leqslant k \leqslant 3,\\
 & \Updownarrow & \\
 \left[ \begin{array}{cccc} b_{k,0} & b_{k,1} & b_{k,2} & b_{k,3} \end{array} \right] &=& %
 T_0[a_{k,0}]\\ %
 & \oplus &
 T_1[a_{(k-1)\bmod N,1}]\\
 & \oplus &
 T_2[a_{(k-2)\bmod N,2}]\\ %
 & \oplus &
 T_3[a_{(k-3)\bmod N,3}],\\
& 0 & \leqslant k \leqslant 3.
\end{eqnarray*}

\subsection{8-bit processors}\label{smart-cards}

On an 8-bit processor with a limited amount of RAM, e.g. a
typical smart card processor, the previous approach is not
feasible. On these processors the substitution is performed byte
by byte, combined with the $\tau$ and the $\sigma[K^r]$
transformation. For $\theta$, it is necessary to implement the
matrix multiplication.

The following piece of pseudo-code calculates one row of $b =
\theta(a)$, using a table $X$ that implements multiplication by
the polynomial $g(x) = x$ in $\GF(2^8)$, i.e. $X[u] = x \cdot u$,
and four registers $r_0, r_1, r_2, r_3$:
\begin{tabbing}margin \= \kill
\> $r_0 = X[a_{i1} \oplus a_{i3}]$;\\
\> $r_1 = X[a_{i0} \oplus a_{i2}]$;\\
\> $r_2 = X[X[a_{i2} \oplus a_{i3}]]$;\\
\> $r_3 = X[X[a_{i0} \oplus a_{i1}]]$;\\
\> $b_{i0} = a_{i0} \oplus r_0 \oplus r_2$;\\
\> $b_{i1} = a_{i1} \oplus r_1 \oplus r_2$;\\
\> $b_{i2} = a_{i2} \oplus r_0 \oplus r_3$;\\
\> $b_{i3} = a_{i3} \oplus r_1 \oplus r_3$;
\end{tabbing}
This implementation requires 12 exors, 6 table lookups and 8
assignments per row. Notice that, if an additional table $X2$ is
available, where $X2[u] \equiv X[X[u]]$, the number of table
lookups drops to 4.

The same $X$ table can be used to implement multiplication by
$\gf{06}$ and $\gf{08}$. The following algorithm computes one row
of $K^r = \phi(\kappa^r)$ using auxiliary registers $u$, $v$, and
$k$:
\begin{tabbing}margin \= \kill
    \> set $K^r_{i0} = K^r_{i1} = K^r_{i2} = K^r_{i3} = S[\kappa^r_{N-1,\,i}]$; \\
    \> for \= ($k = N - 2$; $k \geqslant 0$; $k\!-\!\!-$) \{ \\
    \>     \> set $u = S[\kappa^r_{ki}]$; \\
    \>     \> set $v = X[K^r_{i2}]$; \\
    \>     \> set $K^r_{i0} = u \,\oplus\, K^r_{i0}$; \\
    \>     \> set $K^r_{i1} = u \,\oplus\, X[K^r_{i1}]$; \\
    \>     \> set $K^r_{i2} = u \,\oplus\, X[v] \,\oplus\, v$; \\
    \>     \> set $K^r_{i3} = u \,\oplus\, X[X[X[K^r_{i3}]]]$; \\
    \> \}
\end{tabbing}
This implementation requires $5(N-1)$ exors, $1 + 7(N-1)$ table
lookups and $4 + 6(N-1)$ assignments. With the extra $X2$ table
the number of table lookups drops to $1 + 6(N-1)$ and the number
of assignments to $4 + 5(N-1)$.

\subsection{Techniques to avoid software implementation weaknesses}

The attacks of Kocher \emph{et al.} \cite{timing,power} have raised the
awareness that careless implementation of cryptographic primitives
can be exploited to recover key material. In order to counter this
type of attacks, attention has to be given to the implementation of
the round transformation as well as the key scheduling of the primitive.

A first example is the \emph{timing attack} \cite{timing} that can be applicable
if the execution time of the primitive depends on the value of
the key and the plaintext. This is typically caused by the
presence of conditional execution paths. For instance, multiplication
by a constant value over a finite field is sometimes implemented
as a multiplication followed by a reduction, the latter being
implemented as a conditional exor. This vulnerability is avoided
by implementing the multiplication by a constant by means of table
lookups, as proposed in sections~\ref{large-processors} and~\ref{smart-cards}.

A second class of attacks are the attacks based on the careful
observation of the power consumption pattern of an encryption
device \cite{power}. Protection against this type of attack can
only be achieved by combined measures at the hardware and
software level. We leave the final word on this issue to the
specialists, but we hope that the simple structure and the
limited number of operations in \textsc{\mbox{Anubis}} will make
it easier to create an implementation that resists this type of
attacks.

% can we say more about power attacks
% countermeasures at the software level ?
%
%We probably could if we could define anything
%similar to blinding.  However, blinding in
%general uses algebraic properties of asymmetric
%ciphers.  It is very unlikely that any such
%property exists for Anubis.

\subsection{Hardware implementation}

We have currently no figures on the available performance and
required area or gate count of \textsc{Anubis} in ASIC or FPGA,
nor do we have a description in VHDL. However, we expect that the
results on \textsc{\mbox{Rijndael}}~\cite{hw1,hw2} will carry
over to a large extent.

\section{Efficiency estimates}\label{efficiency}

\subsection{Key setup}

Table~\ref{key-setup} lists the observed key setup efficiency on
a 550 MHz Pentium III platform. The key setup is implemented with
two tables as described in section~\ref{phi-implementation}. The
storage needed for the tables is $2^8 \times 4 \times 2 = 2048$
bytes.
\begin{table}\centering
\caption{Key setup with two auxiliary tables}\label{key-setup}
\begin{tabular}{|c|c|c|} \hline
key size (bits) & cycles (encryption) & cycles (decryption)\\ \hline %
128 &  3352 &  4527 \\ \hline %
160 &  4445 &  5709 \\ \hline %
192 &  6644 &  8008 \\ \hline %
224 &  8129 &  9576 \\ \hline %
256 &  9697 & 11264 \\ \hline %
288 & 11385 & 12931 \\ \hline %
320 & 13475 & 15169 \\ \hline %
\end{tabular}
\end{table}

The increased cost of the decryption key schedule is due to the
application of $\theta$ to $R-1$ round keys. This is done with the
same tables used for encryption and decryption, together with the
$T_4$ table to invert the implicit $\gamma$ transform. The setup of
decryption keys is therefore between $13\%$ and $35\%$
more expensive than the setup of encryption keys.

We point out that the reference implementation used to measure
efficiency is not fully optimised.  By using dedicated key setup
assembler code for each allowed key size (together with the
$N$-table implementation described in
section~\ref{phi-implementation}), we expect a reduction of the
cycle counts by a factor of at least 2.

\subsection{Encryption and decryption}

Since \textsc{Anubis} has involutional structure, encryption and
decryption are equally efficient for any given number of rounds,
which in turn is determined by the key size. Table~\ref{enc-dec}
summarises the observed efficiency on a 550 MHz Pentium III
platform. We use the four-table implementation described in
section~\ref{rho-implementation}.
\begin{table}\centering
\caption{Encryption/decryption efficiency}\label{enc-dec}
\begin{tabular}{|c|c|c|c|} \hline
key size (bits) & cycles per byte & cycles per block & Mbit/s\\ \hline %
128 & 36.8 & 589 & 119.5 \\ \hline %
160 & 39.3 & 628 & 112.1 \\ \hline %
192 & 41.6 & 665 & 105.9 \\ \hline %
224 & 43.8 & 701 & 100.5 \\ \hline %
256 & 46.3 & 740 &  95.1 \\ \hline %
288 & 48.5 & 776 &  90.7 \\ \hline %
320 & 50.8 & 813 &  86.6 \\ \hline %
\end{tabular}
\end{table}

\section{Advantages}\label{advantages}

By design, \textsc{Anubis} is much more scalable than most
modern ciphers, in the sense of being very fast while avoiding
excessive storage space (for both code and tables) and expensive
or unusual instructions built in the processor; this makes it
suitable for a wide variety of platforms.  The same structure also
favours extensively parallel execution of the component mappings,
and its mathematical simplicity tends to make analysis easier.

\subsection{Comparison with other Square ciphers}

In this section, we list the most important differences between
\textsc{Anubis} and two other ciphers developed according to the
Wide Trail strategy, namely, \textsc{\mbox{Square}} and
\textsc{\mbox{Rijndael}}.
\begin{description}
\item[The involutional structure:] The fact that all components
of \textsc{Anubis} are involutions should in principle reduce the code
size or area in software, respectively hardware applications
that implement both encryption and decryption.
\item[The different S-box:] The S-box of \textsc{Anubis} is
generated in a pseudo-random way. The advantage of this lack of
structure is that providing a simple mathematical description seems
more difficult. The polynomial expansion of the S-box is certainly
more involved. The disadvantages are the suboptimal differential
and linear properties, and the more complex hardware implementation.
\item[A more complex key scheduling:]
The advantage is the improved resistance against key
based attacks, in particular the shortcuts for long keys.
The disadvantage is the higher cost:
slower execution, a reduced key agility, larger code or
gate count.
\end{description}

\subsection{Extensions}

The chosen key selection function is based on matrix
$\vdm_N(\gf{01}, \gf{02}, \gf{06}, \gf{08})$, which is MDS for $N
\leqslant 10$. It is possible to extend the key schedule to
support keys up to 512 bits (i.e. $N \leqslant 16$) by
substituting $\vdm_N(\gf{02}, \gf{15}, \gf{1c}, \gf{3c})$ for the
default \textsc{Anubis} matrix. Especially crafted matrices might
be chosen for particular key sizes; for instance, $\vdm_4(\gf{01},
\gf{02}, \gf{05}, \gf{06})$ may have some implementation
advantages for 128-bit keys.

\section{The name}

\textsc{Anubis} was the Egyptian god of embalming and entombment
-- hence, by extension, the god of `encryption'.  The name seems
therefore suitable for a cipher; besides, this choice makes
\textsc{Anubis} the only cipher with an associated curse against
information privacy invaders \verb!:-)!

\begin{thebibliography}{99}

\bibitem{walsh}K.G.~Beauchamp, ``Walsh functions and their
applications,'' Academic Press, 1975.

\bibitem{impossible}E.~Biham, A.~Biryukov, A.~Shamir,
``Cryptanalysis of Skipjack Reduced to 31 Rounds using Impossible
Differentials,'' \emph{Advances in Cryptology, Eurocrypt'99,
LNCS~1592}, J.~Stern, Ed., Springer-Verlag, 1999, pp.~55--64.

\bibitem{biham-keller}E.~Biham and N.~Keller, ``Cryptanalysis of
reduced variants of \textsc{\mbox{Rijndael}},'' submission to the
\emph{Third Advanced Encryption Standard Candidate Conference}.

\bibitem{joan}J.~Daemen, ``Cipher and hash function design
strategies based on linear and differential cryptanalysis,''
\emph{Doctoral Dissertation}, March 1995, K.U.Leuven.

\bibitem{square}J.~Daemen, L.R.~Knudsen and V.~Rijmen,
``The block cipher \textsc{\mbox{Square}},'' \emph{Fast Software
Encryption, LNCS~1267}, E.~Biham, Ed., Springer-Verlag, 1997,
pp.~149--165.

\bibitem{rijndael}J.~Daemen and V.~Rijmen, ``AES proposal:
\textsc{\mbox{Rijndael}},'' AES submission, {\tt
http://www.nist.gov/aes}.

\bibitem{motto}W.V.~Davies, ``Reading the Past -- Egyptian
Hieroglyphics,'' University of California Press/British Museum,
1987.

\bibitem{feistel}H.~Feistel, ``Cryptography and computer privacy,''
\emph{Scientific American}, v.~228, n.~5, 1973, pp.~15--23.

\bibitem{partial-sum}N.~Ferguson, J.~Kelsey, S.~Lucks,
B.~Schneier, M.~Stay, D.~Wagner, and D.~Whiting, ``Improved
Cryptanalysis of \textsc{\mbox{Rijndael}},'' to appear in
\emph{Fast Software Encryption'00}, Springer-Verlag.

\bibitem{gilbert-minier}H.~Gilbert and M.~Minier, ``A collision
attack on 7 rounds of \textsc{\mbox{Rijndael}},'' \emph{Third
Advanced Encryption Standard Candidate Conference}, NIST,
April~2000, pp.~230--241.

\bibitem{hoffman-kunze}K.~Hoffman and R.~Kunze, ``Linear
Algebra (2nd ed.),'' Prentice~Hall, 1971.

\bibitem{hw1} T.~Ichikawa, T.~Kasuya, M.~Matsui,
``Hardware evaluation of the AES finalists,'' presented at the \emph{Third
Advanced Encryption Standard Candidate Conference}.

\bibitem{interpolation}T.~Jakobsen and L.R.~Knudsen,
``The interpolation attack on block ciphers,'' \emph{Fast Software
Encryption, LNCS~1267}, E.~Biham, Ed., Springer-Verlag, 1997,
pp.~28--40.

\bibitem{counterpane-prng}J.~Kelsey, B.~Schneier, D.~Wagner,
C.~Hall, ``Cryptanalytic Attacks on Pseudorandom Number
Generators,'' \emph{Fast Software Encryption, LNCS~1372},
S.~Vaudenay , Ed., Springer-Verlag, 1998, pp.~168--188.

\bibitem{knudsen-trunc}L.R.~Knudsen, ``Truncated and higher order
differentials,'' \emph{Fast Software Encryption, LNCS~1008},
B.~Preneel, Ed., Springer-Verlag, 1995, pp.~196--211.

\bibitem{nonl} L.R.~Knudsen, M.J.B.~Robshaw,
``Non-linear approximations in linear cryptanalysis,''
\emph{Advances in Cryptology,
Eurocrypt'96, LNCS~1070}, U.~Maurer, Ed., Springer-Verlag, 1996,
pp.~224--236.

\bibitem{knudsen-wagner}L.R.~Knudsen and D.~Wagner,
``On The Structure of Skipjack,'' to appear in \emph{Discrete
Applied Mathematics}, special issue on Coding Theory and
Cryptology, C.~Carlet, Ed.

\bibitem{timing}P.C.~Kocher, ``Timing attacks on implementations of {Diffie-Hellman},
{RSA}, {DSS}, and other systems,'' \emph{Advances in Cryptology, Crypto '96, LNCS~1109},
N. Koblitz, Ed., Springer-Verlag, 1996, pp.~104--113.

\bibitem{power} P. Kocher, J. Jaffe, B. Jun, ``Introduction to differential
power analysis and related attacks,'' available from
http://www.cryptography.com/dpa/technical/.

\bibitem{lidl-niederreiter}R.~Lidl and H.~Niederreiter,
``Introduction to finite fields and their applications,''
Cambridge University Press, 1986.

\bibitem{lucks}S.~Lucks, ``Attacking seven rounds of \textsc{\mbox{Rijndael}}
under 192-bit and 256-bit keys,'' \emph{Third Advanced Encryption
Standard Candidate Conference}, NIST, April~2000, pp.~215--229.

\bibitem{macwilliams-sloane}F.J.~MacWilliams and N.J.A.~Sloane,
``The theory of error-correcting codes,'' \emph{North-Holland
Mathematical Library}, vol.~16, 1977.

\bibitem{idea}X.~Lai, J.L.~Massey and S.~Murphy, ``Markov ciphers
and differential cryptanalysis,'' \emph{Advances in Cryptology,
Eurocrypt'91, LNCS~547}, D.W.~Davies, Ed., Springer-Verlag, 1991,
pp.~17--38.

\bibitem{dsa}National Institute of Standards and Technology,
``FIPS 186-2, Digital Signature Standard (DSS),'' January~27,
2000.

%\bibitem{nyberg}K.~Nyberg, ``Differentially uniform mappings for
%cryptography,'' \emph{Advances in Cryptology, Eurocrypt'93,
%LNCS~765}, T.~Helleseth, Ed., Springer-Verlag, 1994, pp.~55--64.

\bibitem{vincent}V.~Rijmen, ``Cryptanalysis and design of
iterated block ciphers,'' \emph{Doctoral Dissertation},
October 1997, K.U.Leuven.

\bibitem{boomerang}D.~Wagner, ``The Boomerang Attack,'' \emph{Fast
Software Encryption, LNCS~1636}, L.~Knudsen, Ed., Springer-Verlag,
1999, pp.~156--170.

\bibitem{hw2} B.~Weeks, M.~Bean, T.~Rozylowicz, C.~Ficke, ``Hardware
performance simulations of round 2 AES algorithms,''
presented at the \emph{Third
Advanced Encryption Standard Candidate Conference}.

\bibitem{youssef-tavares-heys}A.M.~Youssef, S.E.~Tavares, and
H.M.~Heys, ``A New Class of Substitution-Permutation Networks,''
\emph{Workshop on Selected Areas in Cryptography, SAC'96},
Workshop record, 1996, pp.~132--147.

\bibitem{youssef-mister-tavares}A.M.~Youssef, S.~Mister, and
S.E.~Tavares, ``On the Design of Linear Transformations for
Substitution Permutation Encryption Networks,'' \emph{Workshop
on Selected Areas of Cryptography, SAC'97}, Workshop record,
1997, pp.~40--48.

\end{thebibliography}

\appendix

\section{Generation of the {A\normalsize NUBIS} substitution box}

To ensure that the involutional S-box used by \textsc{Anubis} is
generated verifiably pseudo-randomly, the S-box search algorithm
uses a PRNG similar to the one specified for generating
DSA~\cite{dsa} parameters\footnote{It is pointed out
in~\cite{counterpane-prng} that the DSA PRNG is not suitable for
generating pseudo-random bytes that must be kept secret, e.g. for
establishing secret keys. This is not the present case, as the
PRNG is only used to pseudo-randomly generate the \textsc{Anubis}
involutional S-box in a verifiable fashion.}.

\begin{enumerate}
\item The PRNG is keyed by the 20-byte array $k = \textrm{SHA-1}(m)$,
where $m$ is the ASCII-coded string ``\texttt{An offering which
Anubis gives}'' (in Middle Kingdom Egyptian, \emph{\d{h}tp
d\i\!\!'\ 'Inpw}, cf.~\cite{motto}, pp.~14 and~46).
\item The PRNG maintains two 20-byte arrays, a state $s^{(i)}$ and
a buffer $b^{(i)}$, both initialised to zero.  The value of $b^{(i)}$ is
computed from the key $k$ and the state $x^{(i)}$.  Pseudo-random bytes are
taken from $b^{(i)}$ (for implementation convenience, this is done from right
to left, i.e. from index 19 down to index 0). At any moment, the PRNG keeps
track of the number of remaining bytes on $b^{(i)}$ (initially zero).
\item When the number of remaining available bytes on $b^{(i)}$ drops to zero,
$b^{(i)}$ and $x^{(i)}$ are updated as follows:
    \begin{enumerate}
    \item $b^{(i+1)} = \textrm{SHA-1}((k + s^{(i)}) \bmod 2^{160}).$
    \item $s^{(i+1)} = (s^{(i)} + b^{(i)} + 1) \bmod 2^{160}.$
    \end{enumerate}
where the the arrays represent 160-bit integers in little-endian order (i.e.
the value associated to $x^{(i)}$ is $\sum_{t=0}^{19}{256^t \cdot x^{(i)}[t]}$).
\end{enumerate}

The \textsc{Anubis} S-box is pseudo-randomly generated according to
the following algorithm, which makes use of a stream
of pseudo-random bytes produced by the above PRNG:
\begin{enumerate}
\item Generate a list $p$ containing a random permutation
of all 255 nonzero byte values ($p$ is the \emph{difference list}).
\item Mark all entries of $S$ as initially undefined.
\item For each preimage $x$ of the S-box being constructed do the following:
\begin{enumerate}
    \item Sequentially extract from $p$ the first unused difference $d$
        such that entry $S[x \oplus d]$ is undefined; if none is found,
        start over with a new random permutation $p$.
    \item Let $y = x \oplus d$: set $S[x] = y$ and $S[y] = x$
        (thus implicitly marking entry $S[x]$ and $S[y]$ as defined and
        the difference $d$ as used).
    \item Replace the gap left by $d$ in $p$ by the last element of $p$
        (this step is done this way merely to simplify the handling
        of $p$, which may be implemented as an array of bytes).
\end{enumerate}
\end{enumerate}
At the end of this algorithm, $S$ contains a pseudo-random involution
over $\GF(2^8)$ in such a way that, for all $x \in \GF(2^8)$,
$S[x] \neq x$, and the difference value $S[x] \oplus x$ occurs exactly
twice over the table. Now test $S$ for $\delta \leqslant 8$,
$\lambda \leqslant 16$, and $\nu = 7$; if the test fails, start over
with a new random permutation $p$. Repeat the whole process a number
of times (over 600 million in our actual run) and select the generated
$S$ box with the smallest values of $\delta$ and $\lambda$. A final
inspection of the polynomial and rational expansions of $S$ is conducted
to check for any obvious weakness (quite unlikely for pseudo-randomly
generated S-boxes).

\section{The substitution box}\label{sbox-def}

\begin{verse} \tt
const byte sbox[256] = \{ \\
    0xa7, 0xd3, 0xe6, 0x71, 0xd0, 0xac, 0x4d, 0x79,\\
    0x3a, 0xc9, 0x91, 0xfc, 0x1e, 0x47, 0x54, 0xbd,\\
    0x8c, 0xa5, 0x7a, 0xfb, 0x63, 0xb8, 0xdd, 0xd4,\\
    0xe5, 0xb3, 0xc5, 0xbe, 0xa9, 0x88, 0x0c, 0xa2,\\
    0x39, 0xdf, 0x29, 0xda, 0x2b, 0xa8, 0xcb, 0x4c,\\
    0x4b, 0x22, 0xaa, 0x24, 0x41, 0x70, 0xa6, 0xf9,\\
    0x5a, 0xe2, 0xb0, 0x36, 0x7d, 0xe4, 0x33, 0xff,\\
    0x60, 0x20, 0x08, 0x8b, 0x5e, 0xab, 0x7f, 0x78,\\
    0x7c, 0x2c, 0x57, 0xd2, 0xdc, 0x6d, 0x7e, 0x0d,\\
    0x53, 0x94, 0xc3, 0x28, 0x27, 0x06, 0x5f, 0xad,\\
    0x67, 0x5c, 0x55, 0x48, 0x0e, 0x52, 0xea, 0x42,\\
    0x5b, 0x5d, 0x30, 0x58, 0x51, 0x59, 0x3c, 0x4e,\\
    0x38, 0x8a, 0x72, 0x14, 0xe7, 0xc6, 0xde, 0x50,\\
    0x8e, 0x92, 0xd1, 0x77, 0x93, 0x45, 0x9a, 0xce,\\
    0x2d, 0x03, 0x62, 0xb6, 0xb9, 0xbf, 0x96, 0x6b,\\
    0x3f, 0x07, 0x12, 0xae, 0x40, 0x34, 0x46, 0x3e,\\
    0xdb, 0xcf, 0xec, 0xcc, 0xc1, 0xa1, 0xc0, 0xd6,\\
    0x1d, 0xf4, 0x61, 0x3b, 0x10, 0xd8, 0x68, 0xa0,\\
    0xb1, 0x0a, 0x69, 0x6c, 0x49, 0xfa, 0x76, 0xc4,\\
    0x9e, 0x9b, 0x6e, 0x99, 0xc2, 0xb7, 0x98, 0xbc,\\
    0x8f, 0x85, 0x1f, 0xb4, 0xf8, 0x11, 0x2e, 0x00,\\
    0x25, 0x1c, 0x2a, 0x3d, 0x05, 0x4f, 0x7b, 0xb2,\\
    0x32, 0x90, 0xaf, 0x19, 0xa3, 0xf7, 0x73, 0x9d,\\
    0x15, 0x74, 0xee, 0xca, 0x9f, 0x0f, 0x1b, 0x75,\\
    0x86, 0x84, 0x9c, 0x4a, 0x97, 0x1a, 0x65, 0xf6,\\
    0xed, 0x09, 0xbb, 0x26, 0x83, 0xeb, 0x6f, 0x81,\\
    0x04, 0x6a, 0x43, 0x01, 0x17, 0xe1, 0x87, 0xf5,\\
    0x8d, 0xe3, 0x23, 0x80, 0x44, 0x16, 0x66, 0x21,\\
    0xfe, 0xd5, 0x31, 0xd9, 0x35, 0x18, 0x02, 0x64,\\
    0xf2, 0xf1, 0x56, 0xcd, 0x82, 0xc8, 0xba, 0xf0,\\
    0xef, 0xe9, 0xe8, 0xfd, 0x89, 0xd7, 0xc7, 0xb5,\\
    0xa4, 0x2f, 0x95, 0x13, 0x0b, 0xf3, 0xe0, 0x37 \\
\};
\end{verse}

\end{document}
